A Simple Type Theory for Category Theory

### 2022-02-11 paper/lipics-v2021-sample-article.tex
discussion

\section{Related Work}

@@

>>>>

\section{Discussion}

\paragraph{Directed Type Theory}

@@

\paragraph{Semantic extensions}

\paragraph{Extensions}

While we have demonstrated that many common constructions in category
theory can be expressed using the simple combination of homs, tensors
and units, there are several that are out of reach for SCT due to its
restrictive judgmental structure.

First, while we can add products to the calculus, adding in a
\emph{functor category} ${\cat C}^{\cat D}$ would necessitate adding a
syntax for functors of multiple arguments, to allow for the lambda rule:
\[ \inferrule {\alpha:\cat C, \beta:\cat D \vdash e : \cat E} {\alpha:\cat C \vdash \lambda \beta:\cat D. e : \cat E} \]
This would then require also some extension to profunctors for
allowing 0 or more covariant/contravariant dependencies as well.
%
In addition, we could consider extending our semantics with a notion
of \emph{opposite category}, with the property that a covariant
dependence of a profunctor on a variable in a category should be
equivalent to a contravariant dependence on a variable in the opposite
category.
%
This could possibly be done by introducing a modal distinction
$\alpha:^{+}\cat C$ for covariant dependence and $\alpha:^{-}\cat C$
for contravariant dependence, with the opposite category being a
modality that internalizes flipping the variance as a type
constructor.
%
With these two extensions, possibly, the category contexts for
functors and profunctors could be unified as one notion of a sequence
of object variables with variances.
%
With this unification, it would then be simpler to incorporate a third
extension: a universe object, corresponding to an object
$\textrm{Set}$, which would allow us to internalize the profunctor
judgment in the functor judgment, modeling that in classical category
theory, profunctors are the same as certain functors to the category
of Sets.
%
While all of these extensions seem plausible, they all move away to
the simple ordered type theory structure that allowed us to define
units, tensor and hom profunctors so easily. It is unclear how to
define these types in this extended structure. As a simple example, if
profunctors have many variables of mixed variance, rather than exactly
one of each variance, it is unclear what the appropriate
generalization of the transformation context $\Phi$ should be.

An orthogonal extension to these would be to consider adding a
\emph{discretization} modality $|\cat C|$ that gives the discrete
category on the underlying set of objects of $\cat C$. This would give
us the ability to formalize concepts like ``function on objects'' or
``unnatural transformation'' without the need to define internal
categories. This could be modeled by adding a modal distinction
$\alpha:^{\#}\cat C$ for ``crisp'' dependence on an object of $\cat
C$, i.e., dependence merely on the object of $\cat C$ with no
functoriality/naturality conditions. This would then be internalized
by the discretization modality.
